Lost in the woods: shifting habitats can lead phylogeography astray

### 2025-04-20 manuscript/phylogeo.tex
edits

\begin{figure*}[tb]
    \includegraphics*[width=0.99\textwidth]{figures/dispersal_stats.pdf}
    % \includegraphics*[width=0.48\textwidth]{figures/yule_dispersal.pdf}
    \caption{\label{fig:D_and_v}{\bf Empirical summary statistics of $D$ and ``dispersal velocity'' for different sample sizes.}
    While the diffusion constant $D$ (equal to 1 in this example) can be robustly estimated from simulation data of the underlying model, the dispersal rate $v_w$ (weighted) and $v_b$ (by branch) are not well-defined quantities. For Kingman coalescent tree their numerical value depends strongly on sample size and there is no ground truth to compare to. The expected scaling $v_w \sim \sqrt{n}$ of $v$ estimates is indicated by the dashed line.}
\end{figure*}

@@

@@

A common approach in the field is to make models more complex and sample a larger parameter space, for example allow for a broad distribution of diffusion constants in relaxed random walk models.
Such expansion of parameter space and integration over additional phenomenological parameters like $D$ will result in better fits and make the model more permissive to absorb  misspecifications, but does not address the underlying problem.
The computational resources necessary to infer such models increase rapidly with complexity, and it becomes hard to understand how the data inform and constrain the inferences.
In absence of a good understanding of the dispersal properties past how population dynamics in time and space, simple models that allow transparent interpretation of how results are informed by features of the data are preferable to complex and computationally demanding models that nevertheless can fail to faithfully infer history or essential parameters.

>>>>

@@

\begin{figure*}[tb]
    \includegraphics*[width=0.99\textwidth]{figures/dispersal_stats.pdf}
    \caption{\label{fig:D_and_v}{\bf Empirical summary statistics of $D$ and ``dispersal velocity'' for different sample sizes.}
    While the diffusion constant $D$ (equal to 1 in this example) can be robustly estimated from simulation data of the underlying model, the dispersal rate $v_w$ (weighted) and $v_b$ (by branch) are not well-defined quantities. For Kingman coalescent tree their numerical value depends strongly on sample size and there is no ground truth to compare to. The expected scaling $v_w \sim \sqrt{n}$ of $v$ estimates is indicated by the dashed line.}
\end{figure*}

@@

A common approach in the field is to make models more complex and sample a larger parameter space, for example by allowing for a broad distribution of diffusion constants in relaxed random walk models.
Such expansion of parameter space and integration over additional phenomenological parameters like $D$ will result in better fits and make the model more permissive to absorb misspecifications, but does not address the underlying problem.
The computational resources necessary to infer such models increase rapidly with complexity, it becomes harder to understand how the data inform the inferences, and interpretation of inferred parameter values becomes ambiguous as they might have absorbed effects not explicitly modeled.
In absence of a good understanding of the dispersal properties past how population dynamics in time and space that defines model structure, simple approaches that allow transparent interpretation of how results are informed by features of the data are preferable to complex and computationally demanding models that can nevertheless fail to faithfully infer history or essential parameters.
